---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

# NPSETr

The goal of NPSETr is to simplify the calculation of cumulative rates of surface elevation change and vertical accretion from the National Park Services' Surface Elevation Table (SET) data.

This package is based on the [SETr](https://github.com/swmpkim/SETr) package developed by Kim Cressman for the National Estuarine Research Reserve System (<https://nerrssciencecollaborative.org/project/Cressman18>).

## Installation

You can install NPSETr from [GitHub](https://github.com/) with:

``` r
# Set download options and install prerequisite packages
options(download.file.method = "wininet")
install.packages(c("languageserver", "jsonlite", "rlang", "devtools", "remotes", "broom.mixed", "tidyverse"))

# Install NPSutils from github and upgrade all dependencies
devtools::install_github("nationalparkservice/NPSutils", upgrade='all')

# then install NPSETr
devtools::install_github("laura-feher/NPSETr")
```

## Primary Functions

-   [Load SET or MH data](#load-set-or-mh-data)
    -   [From the most recent NPS I&M SET data package]
    -   [From the NPS I&M SET database]
    -   [From a saved file of raw SET or MH data]
-   [Calculate cumulative change](#calculate-cumulative-change)
-   [Plot cumulative change](#plot-cumulative-change)
-   [Calculate linear rates of change](#calculate-linear-rates-of-change)
-   [Plot and compare rates of change](#plot-and-compare-rates-of-change)
-   [Get park-specific SLR data & rates](#get-park-specific-slr-data-and-rates)
-   [Save SET/MH data, cumulative change, or rates of change to file](#save-setmh-data-cumulative-change-or-rates-of-change-to-file)
-   [Save SLR data & rates to file](#save-slr-data-rates-to-file)

## Load SET or MH data {#load-set-or-mh-data}

The package provides 3 options for downloading SET-MH data and loading it into R:

### From the most recent NPS I&M SET data package

Most users from within NPS will want to get data from the most recent version of the NPS I&M SET data package on DataStore. Use `load_set_data()` to get raw surface elevation table data or `load_mh_data()` to get raw marker horizon data. Use the network code (`network_code`) or park code (`park`) arguments to filter the data to a specific I&M network(s) or park(s) using the appropriate 4-letter code.

```{r message = FALSE, warning = FALSE}
library(NPSETr)

# The `load_` functions utilize the `get_data_package` function from the NPSutils package

set_df <- NPSETr::load_set_data(park = "ASIS")

set_df %>%
    dplyr::select(
        network_code,
        park_code,
        site_name,
        station_code,
        event_date_UTC,
        SET_direction,
        pin_position,
        pin_height_mm
    ) %>%
    head()
```

You can also supply a list of multiple parks or networks:

```{r message = FALSE, warning = FALSE}
set_df <- NPSETr::load_set_data(park = c("ASIS", "COLO"))

set_df %>%
    dplyr::select(
        network_code,
        park_code,
        site_name,
        station_code,
        event_date_UTC,
        SET_direction,
        pin_position,
        pin_height_mm
    ) %>%
    head()
```

### From the NPS I&M SET database

For those with access to the backend of the NPS I&M SET database, you can use `load_set_data()` or `load_mh_data()` to pull the latest data from database using the `db_server` argument. Note that you will also need to be connected to the NPS network or on VPN for this to work. Use the network code (`network_code`) or park code (`park`) arguments to filter the data to a specific I&M network(s) or park(s) using the appropriate 4-letter code.

```{r message = FALSE, warning = FALSE}
library(NPSETr)
library(tidyverse)

# Note for those with access to the DB backend: you'll need to replace "example.server" with a valid server path
# set_df <- NPSETr::load_set_data(network_code = "NETN", db_server = "example.server")

set_df %>%
    dplyr::select(
        network_code,
        park_code,
        site_name,
        station_code,
        event_date_UTC,
        SET_direction,
        pin_position,
        pin_height_mm
    ) %>%
    head()
```

### From a saved file of raw SET or MH data

Alternatively, if you have your data saved in a csv, xlsx, or xls file, you can load it into R by supplying the file path to the saved data file. See `? help(calc_change_cumu)` for required columns and data types if supplying your own data.

```{r message = FALSE, warning = FALSE}
set_df <- NPSETr::load_set_data(file_path = here::here("data", "example_sets.csv"))
# equivalent to file_path = "./data/example_sets.csv"

head(set_df)
```

## Calculate cumulative change {#calculate-cumulative-change}

The function `calc_change_cumu()` can be used to calculate station- or site-level cumulative surface elevation change from SET data or vertical accretion from MH data. The resulting data frame has the mean station- or site-level cumulative change for each sampling date. See `? help(calc_change_cumu)` for details on how cumulative change is calculated.

```{r message=FALSE, warning=FALSE}
# station-level cumulative surface elevation change change
example_cumu_station <- NPSETr::calc_change_cumu(example_sets, level = "station")

head(example_cumu_station)
```

```{r message=FALSE, warning=FALSE}
# site-level cumulative vertical accretion
example_cumu_site <- NPSETr::calc_change_cumu(example_mh, level = "site")

head(example_cumu_site)
```

## Plot cumulative change {#plot-cumulative-change}

The function `plot_cumu()` can be used to plot station- or site-level cumulative surface elevation change from SET data or vertical accretion from MH data. You can also plot SET data and MH data on the same plot by supplying data frames to both the `SET_data` and `MH_data` arguments.

```{r message=FALSE, warning=FALSE}
NPSETr::plot_cumu(SET_data = example_sets, level = "station")
```

```{r message=FALSE, warning=FALSE}
NPSETr::plot_cumu(MH_data = example_mh, level = "station")
```

```{r message=FALSE, warning=FALSE}
library(ggh4x)

# Plot SET and MH data for each station together on a single plot:
NPSETr::plot_cumu(SET_data = example_sets, MH_data = example_mh, columns = 2)
```

## Calculate linear rates of change {#calculate-linear-rates-of-change}

The function `calc_linear_rates()` can be used to calculate simple linear rates of change at the station-or site-level in mm/yr.

```{r message=FALSE, warning=FALSE}
station_rates <- NPSETr::calc_linear_rates(example_sets, level = "station")

station_rates %>%
    dplyr::ungroup() %>%
    dplyr::select(park_code, site_name, station_code, rate, rate_se, rate_level)
```

```{r message=FALSE, warning=FALSE}
site_rates <- NPSETr::calc_linear_rates(example_sets, level = "site")

site_rates %>%
    dplyr::ungroup() %>%
    dplyr::select(park_code, site_name, rate, rate_se, rate_level)
```

## Plot and compare rates of change {#plot-and-compare-rates-of-change}

The function `plot_rate_comps()` can be used to create 'tie fighter' plots that are helpful for visually comparing rates of surface elevation change or vertical accretion between stations or sites:

```{r message=FALSE, warning=FALSE}
NPSETr::plot_rate_comps(example_sets, level = "station")
```

Alternatively, if you've already calculated rates of change and want to plot those, you can use the 'rates' argument to supply a data frame of rates that has columns for station IDs, station-level rates, and station rate std errors with one row per station:

```{r message=FALSE, warning=FALSE}
example_rates <- data.frame(
    "station_code" = c("station_1", "station_2", "station_3"),
    "rate" = c(3.2, 4.0, 5.4),
    "rate_se" = c(1, 0.5, 0.25)) %>%
    dplyr::mutate(network_code = "NCBN",
           park_code = "ABCD",
           site_name = "Site A") %>%
    dplyr::group_by(network_code, park_code, site_name, station_code) %>%
    dplyr::select(network_code, park_code, site_name, station_code, rate, rate_se)

NPSETr::plot_rate_comps(rates = example_rates, level = "station")
```

## Get park specific SLR data and rates {#get-park-specific-slr-data-and-rates}

There are 5 functions for getting current/predicted future sea-level data and rates based on data from the NOAA gauge that is closest to each park:

-   `get_sea_level_data()` can be used to download sea-level data from the NOAA tide gauge that is closest to a specific park.
-   `get_long_slr_rate()` gets the current park-specific "long-term" rate of sea-level rise published by NOAA (<https://tidesandcurrents.noaa.gov/sltrends/>).
-   `get_recent_slr_rate()` calculates a recent rate of sea-level for a specified time period. Defaults to the most recent tidal epoch (2001-2019).
-   `get_future_sea_level_data()` gets future predicted sea-level data based on calculations provided in Sweet et al. 2022 for the 5th National Climate Assessment.
-   `get_future_slr_rate()` calculates future predicted sea-level rise rates based on calculations provided in Sweet et al. 2022 for the 5th National Climate Assessment.

```{r message=FALSE, warning=FALSE}
NPSETr::get_sea_level_data("ASIS") %>%
    head(2)

NPSETr::get_long_slr_rate("ASIS") %>%
    head(2)

NPSETr::get_recent_slr_rate("ASIS") %>%
    head(2)

NPSETr::get_future_sea_level_data("ASIS") %>%
    head(2)

NPSETr::get_future_slr_rate("ASIS") %>%
    head(2)
```

## Save SET/MH data, cumulative change, or rates of change to file {#save-setmh-data-cumulative-change-or-rates-of-change-to-file}

The function `write_set_mh_data()` can be used to write data to a csv file. Data can be raw SET/MH produced by `load_set_data()`, cumulative change produced by `calc_change_cumu()`, or rates of change produced by `calc_linear_rates()`.

Note that the name of the saved csv will be the name of the data frame supplied to `data` suffixed with the current date. The csv will be saved to the current working directory by default.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
NPSETr::write_set_mh_data(data = set_df)
```

Use the `dest_folder` argument to save the csv to a specific location. By default, it will not create folders if the folder location doesn't already exist and it will not overwrite files with the same name. Use `create_folders = TRUE` if you want to save to a folder location that doesn't already exist and use `overwrite = TRUE` if you want to overwrite a file with the same name.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
NPSETr::write_set_mh_data(
    data = set_df,
    dest_folder = here::here("C:", "Documents", "data"),
    create_folders = TRUE,
    overwrite = TRUE
)
# this is equivalent to "C:/Documents/SET_data"
```
